---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# indicatorrecode

<!-- badges: start -->
<!-- badges: end -->

The goal of indicatorrecode is to ...


# Context:  *indicator variables well understood*

- https://emilyriederer.netlify.app/post/column-name-contracts/

# Problem:  *content loss using indicator variables directly in data products*

```{r cars}
library(tidyverse)

tidytitanic::passengers %>% 
  ggplot() + 
  aes(x = survived) + 
  geom_bar()

tidytitanic::passengers %>% 
  count(survived)

tidytitanic::passengers %>% 
  janitor::tabyl(sex, survived)

tidytitanic::passengers %>% 
ggplot() + 
  aes(x = sex) + 
  geom_bar() + 
  facet_grid(~ survived)
```


# One-off solution:  *manual indicator recode, but is repetitive*

addition downside is unanticipated reordering depending on value of true case v. false case category 

```{r}
tidytitanic::passengers %>% 
  tibble() %>% 
  mutate(cat_survived = ifelse(survived, "survived", "not survived"), 
         .before = 1)

tidytitanic::passengers %>% 
ggplot() + 
  aes(x = sex) + 
  geom_bar() + 
  facet_grid(~ ifelse(survived, "survived", "not survived"))
```



# Proposed many-times solution:  *ind_recode() function uses variable name as starting point for creating true and false human understandable categories*
  
```{r}
ind_recode <- function(var, var_prefix = "ind_", negator = "not", cat_true = NULL, cat_false = NULL, rev = FALSE){
  
  if(is.null(cat_true)){
    cat_true = deparse(substitute(var)) %>% 
      str_remove(paste0("^", var_prefix)) %>% 
      str_replace_all("_", " ")
  }
  
  if(is.null(cat_false)){
    cat_false = paste(negator, cat_true)
  }
  
  # for yes/no case
  if(is.character({{var}})){
    
    my_var <- {{var}} %>% as.factor() %>% as.numeric() - 1
    
  }else{
      
    my_var <- {{var}}
    }
    
      if(rev){
    ifelse(my_var, cat_true, cat_false) %>% 
      factor(levels = c(cat_true, cat_false))
  }else{
    ifelse(my_var, cat_true, cat_false) %>% 
      factor(levels = c(cat_false, cat_true))
  }
    

}

tibble(ind_female = c(0,0,1,1,1 ,0 ,0)) %>% 
  mutate(cat_female  = ind_recode(ind_female))

tibble(ind_female = c(T,T,F)) %>% 
  mutate(cat_female  = ind_recode(ind_female))

tibble(ind_female = c("Y", "N")) %>% 
  mutate(cat_female  = ind_recode(ind_female))

tibble(ind_female = c("y", "n")) %>% 
  mutate(cat_female  = ind_recode(ind_female))

tibble(ind_female = c("yes", "no")) %>% 
  mutate(cat_female  = ind_recode(ind_female))


```

# Examples:  *How to use ind_recode()*

```{r}
tidytitanic::passengers %>% 
  tibble() %>% 
  mutate(cat_survived = ind_recode(survived))

tidytitanic::passengers %>% 
ggplot() + 
  aes(x = ind_recode(survived)) + 
  geom_bar()

# or
last_plot() +
  aes(x = ind_recode(survived, cat_false = "perished"))

  
# or
last_plot() +
  aes(x = ind_recode(survived, cat_false = "didn't", rev = T)) + 
  labs(x = NULL)

tidytitanic::passengers %>% 
ggplot() + 
  aes(x = sex) + 
  geom_bar() + 
  facet_grid(~ ind_recode(survived))
```

# Known Limitations:  *not for use with magrittr pipe (but base pipe works!)*

```{r}
tidytitanic::passengers %>% 
ggplot() + 
  aes(x = sex) + 
  geom_bar() + 
  facet_grid(~ survived %>% ind_recode())

tidytitanic::passengers %>% 
ggplot() + 
  aes(x = sex) + 
  geom_bar() + 
  facet_grid(~ survived |> ind_recode())
```

# Other questions: *1) Is there already a solution and 2) fundamental problems with this approach?*

Please lemme know 'em! 


# example with tidytuesday data, Spam email

https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-08-15

```{r}
read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-08-15/spam.csv") %>% 
  rename(spam = yesno) %>% 
  ggplot() + 
  aes(fill = ind_recode(bang>0), x = ind_recode(spam)) + 
  geom_bar(position = "dodge")

remove_layers <- function(plot, index = NULL){
  
  if(is.null(index)){
  plot$layers <- NULL
  }else{
  plot$layers[[index]] <- NULL
  }
  
 plot
  
}

read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-08-15/spam.csv") %>% 
  rename(spam = yesno) %>% 
  ggplot() + 
  aes(fill = ind_recode(bang>0), x = ind_recode(spam)) + 
  geom_bar(position = "dodge") -> 
  p; p |> remove_layers() +
  geom_bar(position = "fill")

```
