# Introduction

Recoding indicator variable values to meaningful and appropriately ordered categories often involves redundancy. 

*more description of example here*

```{r, eval = T, message=F, warning=F}
library(tidyverse)

data.frame(ind_graduated = c(T,T,F)) |>
  mutate(cat_graduated  = ifelse(ind_graduated, 
                                 "graduated", 
                                 "not graduated")) |>
  mutate(cat_graduated = fct_rev(cat_graduated))  
```

ind2cat's ind_recode function avoids repetition by creating categories based on the indicator variable name. Using the the function ind_recode(), we can accomplish the same task shown above more succinctly:

```{r, eval = T, message=F, warning=F}
library(ind2cat)

data.frame(ind_graduated = c(T,T,F)) |>
  mutate(cat_graduated  = ind_recode(ind_graduated))
```

Furthermore, ind_recode's functionality allows analysts to move from a first-cut recode that delivers meaningful categories to fully customized categories.


```{r}
data.frame(ind_graduated = c(T,T,F)) %>% 
  mutate(cat_graduated  = ind_recode(ind_graduated, 
                                     cat_false = "current"))
```



# Status-Quo: Analysts make a choice between directly using indicators or verbose recode

## Current proceedures for recoding indicator variables to a categorial variable is inelegant.

Current methods for recoding indicator variables is repetitive and verbose, as shown in the examples that follow. 

```{r}
tidytitanic::passengers %>% 
  tibble() %>% 
  mutate(cat_survived = ifelse(survived, "survived", "not survived"), 
         .before = 1)

tidytitanic::passengers %>% 
ggplot() + 
  aes(x = sex) + 
  geom_bar() + 
  facet_grid(~ ifelse(survived, "survived", "not survived"))
```

This solution above also does not address category display ordering; ordering in products will be alphabetical and not reflect the F/T order of the source variable.  An additional step to reflect the source variable, using a function like forcats::fct_rev, may be required for consistency in reporting.

```{r}
data.frame(ind_daytime = c(T, F, T, T)) %>% 
    mutate(cat_survived = ifelse(ind_daytime, "daytime", "not daytime")) %>% 
  mutate(cat_survived = fct_rev(cat_survived)) %>% 
  ggplot() + 
  aes(x = cat_survived) + 
  geom_bar()
```

## Direct use of indicator variables in data products makes product more difficult or impossible to interpret.

Given how verbose recoding can be, analyst may choose to forego a recoding the variable, especially in exploratory analysis.

When indicator variables are not translated to a categorical analogue in creating data products like tables and visuals, information is often awkwardly displayed and is sometimes lost. 
When creating tables, using an indicator variable directly can be awkward or insufficient for interpretation.

Below, the column header from variable name and 0-1 categories preserves information but is awkward:

```{r}
tidytitanic::passengers %>% 
  count(survived) 
```

In the following two-way table, information is completely lost:

```{r}
tidytitanic::passengers %>% 
  janitor::tabyl(sex, survived) %>% 
  knitr::kable(caption = "C. ", format = kabel_format)
```

Furthermore in creating ba

```{r cars, fig.cap="A. Bar labels + axis label preserves information but is awkward"}
library(tidyverse)

tidytitanic::passengers %>% 
  ggplot() + 
  aes(x = survived) + 
  geom_bar()
```

```{r, fig.cap = "D. Facetting directly on an indicator variable with popular ggplot2 results in information loss"}
tidytitanic::passengers %>% 
ggplot() + 
  aes(x = sex) + 
  geom_bar() + 
  facet_grid(~ survived)
```



# Introducing ind_recode  *ind_recode() function uses variable name as starting point for human-readable categories*
  
<!-- Usually I'll start with the sketch of a function right here, but then I eventually move it to an .R file. -->
  
```{r, echo = F}
code = readLines("R/ind_recode.R")
```
  
```{r, code = code, eval = T}

```

# to do

- change to Rlang for grabbing function name
- make "Y" "N" a lot stricter - right now we're assuming a ton! Eek!

# Basic examples:  *How to use ind_recode()* 

```{r}
library(tibble)
tibble(ind_grad = c(0,0,1,1,1 ,0 ,0)) %>%
  mutate(cat_grad  = ind_recode(ind_grad))

tibble(ind_grad = c(T,T,F)) %>%
  mutate(cat_grad  = ind_recode(ind_grad))

tibble(ind_grad = c("Y", "N")) %>%
  mutate(cat_grad  = ind_recode(ind_grad))

tibble(ind_grad = c("y", "n")) %>%
  mutate(cat_grad  = ind_recode(ind_grad))

tibble(ind_grad = c("yes", "no")) %>%
  mutate(cat_grad  = ind_recode(ind_grad))

```


# Customizability

We believe that ind_recode is useful in quickly translating to a human understandable outcome.  

However, addition functionality allows analysts to fully specify their preferences about the categories outputted.  


- var_prefix a character string that will be ignored when creating the categorical variable
- negator a character string used to create cat_false when cat_false is NULL, default is 'not'
- cat_true a character string string to be used place of  T/1/"Yes" for the categorical variable output, if NULL the category is automatically generated from the variable name
- cat_false a character string string to be used place of  F/0/"No" for the categorical variable output, if NULL the category is automatically generated from cat_true and the negator
- rev logical indicating if the order should be reversed from the F/T ordering of the indicator source variable, default is FALSE



## Customization examples


```{r}
tibble(dummy_grad = c(0,0,1,1,1 ,0 ,0)) %>%
  mutate(cat_grad  = ind_recode(dummy_grad, var_prefix = "dummy_"))

tibble(ind_grad = c(T,T,F)) %>%
  mutate(cat_grad  = ind_recode(ind_grad, negator = "didn't"))

tibble(ind_grad = c("Y", "N")) %>%
  mutate(cat_grad  = ind_recode(ind_grad, cat_false = "enrolled"))

tibble(ind_grad = c("y", "n")) %>%
  mutate(cat_grad  = ind_recode(ind_grad, 
                                cat_true = "graduated"))

tibble(ind_grad = c("y", "n")) %>%
  mutate(cat_grad  = ind_recode(ind_grad, 
                                cat_true = "graduated", 
                                cat_false = "enrolled"))

tibble(ind_grad = c("yes", "no")) %>%
  mutate(cat_grad  = ind_recode(ind_grad, rev = TRUE)) %>% 
  mutate(cat_grad_num = as.numeric(cat_grad))
```

## Use in data products like figures and tables



```{r}
tidytitanic::passengers %>% 
ggplot() + 
  aes(x = ind_recode(survived)) + 
  geom_bar()

# or
last_plot() +
  aes(x = ind_recode(survived, cat_false = "perished"))

  
# or
last_plot() +
  aes(x = ind_recode(survived, cat_false = "didn't", rev = T)) + 
  labs(x = NULL)

tidytitanic::passengers %>% 
ggplot() + 
  aes(x = sex) + 
  geom_bar() + 
  facet_grid(~ ind_recode(survived))

```

# Known Limitations:  *not for use with magrittr pipe (but base pipe works!)*

```{r}
tidytitanic::passengers %>% 
ggplot() + 
  aes(x = sex) + 
  geom_bar() + 
  facet_grid(~ survived %>% ind_recode())

tidytitanic::passengers %>% 
ggplot() + 
  aes(x = sex) + 
  geom_bar() + 
  facet_grid(~ survived |> ind_recode())
```


---

Afterward

# Other questions: *1) Is there already a solution and 2) fundamental problems with this approach?*

Please lemme know 'em! 

---

# worked example with tidytuesday data, Spam email

https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-08-15

```{r}
read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-08-15/spam.csv") %>% 
  rename(spam = yesno) %>% 
  ggplot() + 
  aes(fill = ind_recode(bang>0), x = ind_recode(spam)) + 
  geom_bar(position = "dodge")

remove_layers <- function(plot, index = NULL){
  
  if(is.null(index)){
  plot$layers <- NULL
  }else{
  plot$layers[[index]] <- NULL
  }
  
 plot
  
}

last_plot_wiped <- function(index = NULL){
  
  plot <- last_plot()
  
  if(is.null(index)){
  plot$layers <- NULL
  }else{
  plot$layers[[index]] <- NULL
  }
  
 plot
  
}

last_plot_wiped() +
  geom_bar(position = "fill")


```
# learned along the way: `as_factor()` has different behavior than `as.factor()`

```{r}
c("Y", "N") %>% as_factor()
c("Y", "N") %>% as.factor()
```
