---
title: "introducing {{ind2cat}}"
subtitle: "Easily recode indicator variables to categories"
author: "Gina Reynolds"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, hygge, ninjutsu]
    nature:
      ratio: 16:9
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r, include = F}
# This is the recommended set up for flipbooks
# you might think about setting cache to TRUE as you gain practice --- building flipbooks from scratch can be time consuming
knitr::opts_chunk$set(fig.width = 6, message = FALSE, warning = FALSE, comment = "", cache = T)
library(flipbookr)
library(tidyverse)
```


```{r, child="../README.Rmd", include=F}

```

```{r}
ggplot2::theme_set(new = theme_gray(base_size = 25))

```


<!-- get a flipbook version of the my_cars code chunk - pauses are where parentheses are balanced -->




---
class: inverse

<br><br><br><br>

# Indicator variables are easy to create, store and interpret.

--

# But direct use of indicator variables to create some data products yields bad results.

---

## Titanic example


```{r example_data}
library(tidytitanic)
passengers %>% 
  mutate(name = str_extract(name, ".{18}")) %>% 
  head() 
```



---

`r chunk_reveal("direct_table_awkward", break_type = 1, title = "## Awkward table output", widths = c(1,1))`

---

`r chunk_reveal("direct_visual_awkward", break_type = 1, title = "## Awkward visualization output", widths = c(1,1))`

---

`r chunk_reveal("direct_table_loss", break_type = 1, title = "## Info loss in table output", widths = c(1,1))`

---

`r chunk_reveal("direct_visual_loss", break_type = 1, title = "## Info loss in visual output", widths = c(1,1))`

---
class: inverse

<br><br><br><br>


# Recoding to a categorical variable yeilds better products.

--

# But is recoding is currently repetative.

---

`r chunk_reveal("manipulation_status_quo_reprise", title = "## Status quo pipeline", widths = c(1,1))`

---


# introducing ind2cat::ind_recode()

---


`r chunk_reveal("manipulation_ind2cat", title = "## Proposed ind2cat pipeline", widths = c(1,1))`


---



---





<!-- adjust font size in this css code chunk, currently 80 -->

```{css, eval = TRUE, echo = FALSE}
.remark-code{line-height: 1.5; font-size: 130%}

@media print {
  .has-continuation {
    display: block;
  }
}

code.r.hljs.remark-code{
  position: relative;
  overflow-x: hidden;
}


code.r.hljs.remark-code:hover{
  overflow-x:visible;
  width: 500px;
  border-style: solid;
}
```

